---
title: "General profile of people voted"
output: html_notebook
---

### Introduction
This is a data story about the general profile of people who participated in the voteoted. I'm curious about things like gender, education level, age of people who participated in vote. Will the profile of participated people change over year? Or will it remain stable?

```{r}
# import data
library(readstata13)
mydata<-read.dta13('D:\\Applied data\\anes_timeseries_cdf_dta\\anes_timeseries_cdf.dta')
mydata<-data.frame(lapply(mydata, as.character), stringsAsFactors=FALSE)
```

```{r}
# check what data looks like
print(dim(mydata)[1])
print(dim(mydata)[2])
head(mydata, 5)
```

We need to extract some useful columns from this big dataset. I extracted five columns----'year', 'age', 'gender', 'race' and 'education'. In addition, I only use data after 1984, because questions in census changed frequently before 1984. Besides, there will be too many data if we use all the data since 1948.

```{r}
# find useful variables
data<-mydata[,c('VCF0004', 'VCF0102', 'VCF0104', 'VCF0105b', 'VCF0110')]
data<-data[which(data$VCF0004>=1984),]
print(dim(data)[1])
print(dim(data)[2])
head(data, 5)
```

Now we have a dataframe of data we needed, we need to do some simple check about whether there are problems inside data. 

```{r}
# check data status
MissingValue<-c()
UniqueValue<-c()
MostFreqValue<-c()
MostFreqValueRate<-c()
for (i in colnames(data)){
  MissingValue<-c(MissingValue, sum(is.na(data[,i])))
  UniqueValue<-c(UniqueValue, round(length(unique(data[,i])),2))
  temp<-sort(table(data[,i], useNA = "ifany"), decreasing = TRUE)
  MostFreqValue<-c(MostFreqValue, names(temp)[1])
  MostFreqValueRate<-c(MostFreqValueRate, round(temp[1]/nrow(data),2))
}
df<-data.frame(MissingValue, UniqueValue, MostFreqValue, MostFreqValueRate)
rownames(df)<-c('year', 'age', 'gender', 'race', 'education')
df
```

The dataset has few missing value, and there are not many repeated value in the dataset. Since there are not many missing value, I simply delete all the missing value.

```{r}
# delete na data
newdata<-na.omit(data)
colnames(newdata)<-c('year', 'age', 'gender', 'race', 'education')
print(dim(newdata)[1])
print(dim(newdata)[2])
```

I used percentage histogram to visualize each variables.

```{r}
library(dplyr)
library(plyr)
library(ggplot2)
# gender
a<-newdata %>%
  group_by(year, gender) %>%
  dplyr::summarize(n=n())
ce<-ddply(a, "year", transform, percent_weight = n / sum(n) * 100)
ggplot(ce, aes(x = year, y = percent_weight, fill = gender)) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_brewer(palette = "Pastel1")+
  geom_text(aes(label = round(percent_weight,2), y = percent_weight), size = 2, 
            position = position_stack(vjust = 0.5))+
  theme(axis.text.x = element_text(angle = 60))
```
The porpotion of male and female is stable over year. Female is slightly fewer male.

```{r}
# race
a<-newdata %>%
  group_by(year, race) %>%
  dplyr::summarise(n=n())
ce<-ddply(a, "year", transform, percent_weight = n / sum(n) * 100)
ggplot(ce, aes(x = year, y = percent_weight, fill = race)) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_brewer(palette = "Pastel1")+
  geom_text(aes(label = round(percent_weight,2), y = percent_weight), size = 2, 
            position = position_stack(vjust = 0.5))+
  theme(axis.text.x = element_text(angle = 60))
```
The porpotion of different race of people participated in vote is stable until 2000. Maybe this is because survey sample change. 

```{r}
# age
a<-newdata %>%
  group_by(year, age) %>%
  dplyr::summarise(n=n())
ce<-ddply(a, "year", transform, percent_weight = n / sum(n) * 100)
ggplot(ce, aes(x = year, y = percent_weight, fill = age)) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_brewer(palette = "Pastel1")+
  geom_text(aes(label = round(percent_weight,2), y = percent_weight), size = 2, 
            position = position_stack(vjust = 0.5))+
  theme(axis.text.x = element_text(angle = 60))
```
The porpotion of different age of people participated in vote is not stable. From the graph we can see number of people in age group 1 going down first then going up and then going down. People in age group 2 keeping going down. People in age group 3 and 4 going up then going down. people in age group 5 keep going up. In order to visualize the pattern, I will produce a line chart later.

```{r}
# education
a<-newdata %>%
  group_by(year, education) %>%
  dplyr::summarise(n=n())
ce<-ddply(a, "year", transform, percent_weight = n / sum(n) * 100)
ggplot(ce, aes(x = year, y = percent_weight, fill = education)) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_brewer(palette = "Pastel1")+
  geom_text(aes(label = round(percent_weight,2), y = percent_weight), size = 2, 
            position = position_stack(vjust = 0.5))+
  theme(axis.text.x = element_text(angle = 60))
```
The porpotion of different education level of people participated in vote is not stable. From the graph we can see number of people in group 1 and 2 keep going down except for 2008. People in group 3 and 4 keep going up except for 2008. I think 2008 year should be exclude from analysis.

```{r}
# line chart for age
a<-newdata %>%
  group_by(year, age) %>%
  dplyr::summarise(n=n())
ce<-ddply(a, "year", transform, percent_weight = n / sum(n) * 100)
temp<-ce[which(ce$age != '6. 65 - 74'),]
temp<-temp[which(temp$age != '7. 75 - 99 and over (except 1954)'),]
ggplot(temp, aes(x=year, y = percent_weight, colour = age, group = age)) +
  geom_line()+
  geom_point()+
  theme(axis.text.x = element_text(angle = 60))
```
From the graph we can see that group 2 number of people participated in vote peak at approximately year 1984. Group 3 peak at approximately 1996. Group 4 peak at approximately 2004, group 5 at 2012. The peak point of each groups are approximately 10 year apart, which is exactly the age difference between each groups. Maybe only the same group of people are willing to vote, young people are not very willing to participate. Age of people participate in vote tend to become elder and elder.

```{r}
# line chart for education
a<-newdata %>%
  group_by(year, education) %>%
  dplyr::summarise(n=n())
ce<-ddply(a, "year", transform, percent_weight = n / sum(n) * 100)
ggplot(ce, aes(x = year, y = percent_weight, colour = education, group=education)) +
  geom_point()+
  geom_line()+
  theme(axis.text.x = element_text(angle = 60))
```
From line chart above we can clearly see that number of people in group 1 and 2 keep decreasing, number of people from group 3 and 4 keep increasing. More and more people with higher level of education participate in vote.

### Summary
White non Hispanic and male people are main voter. Age of people participate in vote tend to become elder and elder. Educational level of people participate in vote tend to become higher and higher.